\modLP{Identification of} Biological Regulatory Networks from Process Hitting models

### 2014-10-30 parts-grn.tex
begin switching from interval parameters to integer parameters in the running example

\begin{example}
\pref{fig:runningBRN}(left) represents an Interaction Graph $(\Gamma,E)$ where
$\Gamma = \{a, b, c\}$, with $l_a = 2$ and $l_b = l_c = 1$,
and:
\begin{align*}
  E_+ &= \{\GRNedgef{b}{+}{1}{a}, \GRNedgef{c}{+}{1}{a}\} &
  E_\uns &= \emptyset \\
  E_- &= \{\GRNedgef{a}{-}{2}{b}\}
\end{align*}
Hence:
\begin{align*}
  \GRNreg{a} &= \{ b, c \} &
  \GRNreg{b} &= \{ a \} \\
  \GRNreg{c} &= \emptyset
\end{align*}
We also have especially:
\begin{align*}
  \levels{a}{b} &= \segm{2}{2} & \ulevels{a}{b} &= \segm{0}{1}
\end{align*}
\end{example}

\begin{figure}[t]
\begin{minipage}{0.4\linewidth}
\centering
\scalebox{1.2}{
\begin{tikzpicture}[grn]
  \path[use as bounding box] (-0.3,-0.75) rectangle (2.5,1.5);
  \node[inner sep=0] (a) at (2,0) {a};
  \node[inner sep=0] (b) at (0,0) {b};
  \node[inner sep=0] (c) at (2,1.2) {c};
  \path[->]
    (b) edge[bend right] node[elabel, below=-2pt] {$+1$} (a)
    (c) edge node[elabel, right=-2pt] {$+1$} (a)
    (a) edge[bend right] node[elabel, above=-5pt] {$-2$} (b);
\end{tikzpicture}
}
\end{minipage}
\begin{minipage}{0.6\linewidth}
\centering
\begin{align*}
  K_{a,\emptyset} &= \segm{0}{0} & K_{b,\emptyset} &= \segm{0}{1} \\
  K_{a,\{b\}} &= \segm{1}{1} & K_{b,\{a\}} &= \segm{0}{0} \\
  K_{a,\{c\}} &= \segm{1}{1} &&\\
  K_{a,\{b,c\}} &= \segm{2}{2} & K_{c,\emptyset} &= \segm{0}{1}
\end{align*}
\end{minipage}
\caption{\label{fig:runningBRN}
  (left)
  IG example.
  Components are represented by nodes labeled with a name
  and regulations by edges labeled with their sign and threshold.
  For instance, the edge from $b$ to $a$ is labeled $+1$, which stands for: $\GRNedgef{b}{+}{1}{a}$.
  This means that if the expression level of $b$ is equal to (\ie above) 1, then $b$ activates $a$;
  otherwise, $b$ inhibits $a$.
  (right)
  Example parametrization on the left IG.
}
\end{figure}

@@

\begin{example}
\pref{fig:runningBRN}(right) gives a Parametrization on the IG of \pref{fig:runningBRN}(left).
In this BRN, the following transitions are possible given the semantics defined in \pref{def:dynamics}:
$$\GRNstate{a_0, b_1, c_1} \rightarrow \GRNstate{a_1, b_1, c_1} \rightarrow \GRNstate{a_2, b_1, c_1} \rightarrow
\GRNstate{a_2, b_0, c_1} \rightarrow \GRNstate{a_1, b_0, c_1} \enspace,$$
where $a_i$ denotes the component $a$ at level $i$.
This sequence of states ends in a steady state: no evolution is possible in $\GRNstate{a_1, b_0, c_1}$.
Due to the semantics of interval parameters and as $K_{b,\emptyset} = \segm{0}{1}$, no auto-regulation on $b$ is needed to prevent its evolution when $a$ is not at level $2$.
\todo{Remove interval parameter}
\todo{Change example to preserve the dynamics?}
\end{example}

>>>>

\begin{example}
\pref{fig:runningBRN}(left) represents an Interaction Graph $(\Gamma,E)$ where
$\Gamma = \{a, b, c\}$, with $l_a = 2$ and $l_b = l_c = 1$,
and:
\begin{align*}
  E_+ &= \{\GRNedgef{b}{+}{1}{a}, \GRNedgef{c}{+}{1}{a},
    \GRNedgef{b}{+}{1}{b}, \GRNedgef{c}{+}{1}{c}\} \\
  E_- &= \{\GRNedgef{a}{-}{2}{b}\} &&
    E_\uns = \emptyset \\
\end{align*}
Hence:
\begin{align*}
  \GRNreg{a} &= \{ b, c \} &
  \GRNreg{b} &= \{ a, b \} \\
  \GRNreg{c} &= \{ c \}
\end{align*}
We also have especially:
\begin{align*}
  \levels{a}{b} &= \segm{2}{2} & \ulevels{a}{b} &= \segm{0}{1}
\end{align*}
\end{example}

\begin{figure}[t]
\begin{minipage}{0.4\linewidth}
\centering
\scalebox{1.2}{
\begin{tikzpicture}[grn]
  \path[use as bounding box] (-0.5,-0.75) rectangle (3.2,1.7);
  \node[inner sep=0] (a) at (2,0) {a};
  \node[inner sep=0] (b) at (0,0) {b};
  \node[inner sep=0] (c) at (2,1.2) {c};
  \path[->]
    (b) edge[bend right] node[elabel, below=-2pt] {$+1$} (a)
    (c) edge node[elabel, right=-2pt] {$+1$} (a)
    (a) edge[bend right] node[elabel, above=-5pt] {$-2$} (b)
    (b) edge[loop, min distance=30pt] node[elabel, above=-5pt] {$+1$} (b)
    (c) edge[loop right, min distance=30pt] node[elabel, above=0pt, xshift=-10] {$+1$} (c);
\end{tikzpicture}
}
\end{minipage}
\begin{minipage}{0.6\linewidth}
\centering
\modMF{
\begin{align*}
  K_{a,\emptyset} &= 0 & K_{b,\emptyset} &= 0 \\
  K_{a,\{b\}} &= 1 & K_{b,\{a\}} &= 0 \\
  K_{a,\{c\}} &= 1 & K_{b,\{b\}} &= 1 \\
  K_{a,\{b,c\}} &= 2  & K_{b,\{a,b\}} &= 0 \\
  \\
  K_{c,\emptyset} &= 0 & K_{c,\{c\}} &= 1
\end{align*}
}
\end{minipage}
\caption{\label{fig:runningBRN}
  (left)
  IG example.
  Components are represented by nodes labeled with a name
  and regulations by edges labeled with their sign and threshold.
  For instance, the edge from $b$ to $a$ is labeled $+1$, which stands for: $\GRNedgef{b}{+}{1}{a}$.
  This means that if the expression level of $b$ is equal to (\ie above) 1, then $b$ activates $a$;
  otherwise, $b$ inhibits $a$.
  (right)
  Example parametrization on the left IG.
}
\end{figure}

@@

\begin{example}
\pref{fig:runningBRN}(right) gives a Parametrization on the IG of \pref{fig:runningBRN}(left).
In this BRN, the following transitions are possible given the semantics defined in \pref{def:dynamics}:
$$\GRNstate{a_0, b_1, c_1} \rightarrow \GRNstate{a_1, b_1, c_1} \rightarrow \GRNstate{a_2, b_1, c_1} \rightarrow
\GRNstate{a_2, b_0, c_1} \rightarrow \GRNstate{a_1, b_0, c_1} \enspace,$$
where $a_i$ denotes the component $a$ at level $i$.
This sequence of states ends in a steady state: no evolution is possible in $\GRNstate{a_1, b_0, c_1}$.
\end{example}

### 2014-10-30 parts-infer-ig.tex
begin switching from interval parameters to integer parameters in the running example

The IG inference of the PH of \pref{fig:runningPH-2} gives the
IG in \pref{fig:BRN-inf1}, containing the following edges:
\begin{align*}
  E_+ &= \{\GRNedgef{b}{+}{1}{a}, \GRNedgef{c}{+}{1}{a}, \GRNedgef{a}{+}{1}{a}, \GRNedgef{b}{+}{1}{b}, \GRNedgef{c}{+}{1}{c}\}\\
  E_- &= \{\GRNedgef{a}{-}{2}{b}\} \qquad\qquad\qquad\qquad\qquad
  E_\uns = \emptyset
\end{align*}
This IG is close to the one in \pref{fig:runningBRN} but not equivalent,
as each component has an additional auto-action.
The auto-actions on $b$ and $c$ are the consequence of a global stability
in some configurations: $c$ never evolves, and neither does $b$ when $a_2$ is not active.
The auto-action on $a$ is mainly caused by its multi-valued nature.

The inference of the PH of \pref{fig:runningPH-1}
(without refinement with cooperative sort) gives the same IG.

\begin{figure}[t]
\centering
\scalebox{1.2}{
\begin{tikzpicture}[grn]
  \path[use as bounding box] (-1.3,-0.75) rectangle (3.5,1.5);
  \node[inner sep=0] (a) at (2,0) {a};
  \node[inner sep=0] (b) at (0,0) {b};
  \node[inner sep=0] (c) at (2,1.2) {c};
  \path[->]
    (b) edge[bend right] node[elabel, below=-2pt] {$+1$} (a)
    (c) edge node[elabel, right=-2pt] {$+1$} (a)
    (a) edge[bend right] node[elabel, above=-5pt] {$-2$} (b)
    (b) edge[in=-15+180, out=15+180, loop] node[elabel, left=-2pt] {+1} (b)
    (c) edge[in=15, out=-15, loop] node[elabel, right=-2pt] {+1} (c)
    (a) edge[in=15, out=-15, loop] node[elabel, right=-2pt] {+1} (a);
\end{tikzpicture}
}
\caption{\label{fig:BRN-inf1}
  Result of the IG inference performed on the PH of \pref{fig:runningPH-2}.
}
\end{figure}

If we add the action $\PHfrappe{a_2}{b_0}{b_1}$ to the PH of \pref{fig:runningPH-2},
then two unsigned edges towards $b$ are inferred instead of the previous signed edges:
\begin{align*}
  E_+ &= \{\GRNedgef{b}{+}{1}{a}, \GRNedgef{c}{+}{1}{a}, \GRNedgef{a}{+}{1}{a}, \GRNedgef{c}{+}{1}{c}\}\\
  E_- &= \emptyset \qquad\qquad\qquad\qquad
  E_\uns = \{\GRNedgef{a}{\uns}{2}{b}, \GRNedgef{b}{\uns}{1}{b}\}
\end{align*}
This is due to the fact that the actions $\PHfrappe{a_2}{b_1}{b_0}$ and $\PHfrappe{a_2}{b_0}{b_1}$
introduce an oscillation only caused by $a$, which cannot be represented in Thomas modeling.
\todo{Refresh this example for the new inference}

>>>>

\modMF{
The IG inference of the PH of \pref{fig:runningPH-2} gives the
IG in \pref{fig:runningBRN}(left), that is, allows to find back all the edges
and their signs and thresholds.
This is mainly due to the fact that the cooperation between $b$ and $c$ is well-defined,
using a (deterministic) cooperative sort.
}

\modMF{
Indeed, the inference of the PH of \pref{fig:runningPH-1},
that is, without refinement with a cooperative sort, gives the following IG:
\begin{align*}
  E_+ &= \{\GRNedgef{b}{+}{1}{a}, \GRNedgef{c}{+}{1}{a}, \GRNedgef{c}{+}{1}{c}\}\\
  E_- &= \{\GRNedgef{a}{-}{2}{b}\} \qquad\qquad\qquad\qquad\qquad
  E_\uns = \emptyset
\end{align*}
This IG is close to the one in \pref{fig:runningBRN} but misses the self-influence
$\GRNedgef{b}{+}{1}{b}$.
\todo{Detail why?}
}

\modMF{
If we add the action $\PHfrappe{a_2}{b_0}{b_1}$ to the PH of \pref{fig:runningPH-2},
then two unsigned edges towards $b$ are inferred instead of the previous signed edges:
\begin{align*}
  E_+ &= \{\GRNedgef{b}{+}{1}{a}, \GRNedgef{c}{+}{1}{a}, \GRNedgef{a}{+}{1}{a}, \GRNedgef{c}{+}{1}{c}\}\\
  E_- &= \emptyset \qquad\qquad\qquad\qquad
  E_\uns = \{\GRNedgef{a}{\uns}{2}{b}, \GRNedgef{b}{\uns}{1}{b}\}
\end{align*}
This is due to the fact that the actions $\PHfrappe{a_2}{b_1}{b_0}$ and $\PHfrappe{a_2}{b_0}{b_1}$
introduce an oscillation only caused by $a$, which cannot be represented in Thomas modeling.
}

### 2014-10-30 parts-infer-param.tex
begin switching from interval parameters to integer parameters in the running example

\begin{example}
\label{ex:infer-param-runningPH-1}

Applied to the PH in \pref{fig:runningPH-1}, we infer the following parameters:
\begin{align*}
K_{a, \emptyset} &= 0&
K_{b, \emptyset} &= 0 \\
K_{a, \{a\}} &= 0 &
K_{b, \{a\}} &= 0 \\
K_{a, \{c\}} &= 1 &
K_{b, \{b\}} &= 1 \\
K_{a, \{b\}} &= 1 &
K_{b, \{a,b\}} &= 0 \\
K_{a, \{b,c\}} &= 1 &
K_{c, \emptyset} &= 0 \\
K_{a, \{a,b,c\}} &= 2 &
K_{c, \{c\}} &= 1
%K_{a, \{a,b\}} = ?
%K_{a, \{a,c\}} = ?
\end{align*}
$K_{a,\{a,b\}}$ and $K_{a,\{a,c\}}$ cannot be inferred,
which is a direct consequence of the lack of cooperation between $b$ and $c$ on $a$.
\end{example}

\begin{example}
Regarding the refined PH of \pref{fig:runningPH-2}, all parameters can be inferred.
We obtain the same value for the inferred parameters of \pref{ex:infer-param-runningPH-1},
together with the following results:
\begin{align*}
  K_{a,\{a,b\}} &= 1 &
  K_{a,\{a,c\}} &= 1
\end{align*}
\end{example}

>>>>

\begin{example}
\label{ex:infer-param-runningPH-1}
When applied to the refined PH in \pref{fig:runningPH-2},
and with the IG of \pref{fig:runningBRN}(left)
(which is the result of the IG inference on this PH model,
as explained in \pref{ssec:infer-ig-examples}),
\label{pps:param_K} infers the parametrization of \pref{fig:runningBRN}(right).
In other words, the inference of parameters allows to find back all parameters
of the original model when the cooperations are fully defined.
\end{example}

$K_{a,\{a,b\}}$ and $K_{a,\{a,c\}}$ cannot be inferred,
which is a direct consequence of the lack of cooperation between $b$ and $c$ on $a$.

\begin{example}
Regarding the non-refined PH of \pref{fig:runningPH-1},
with the IG produced by the IG inference described in \pref{ssec:infer-ig-examples}
(that is, the same as \pref{fig:runningBRN}(left)
without the self-influence $\GRNedgef{b}{+}{1}{b}$)
the following parameters are inferred:
\begin{align*}
  K_{a,\emptyset} &= 0 &
  K_{b,\{a\}} &= 0 \\
%  K_{a,\{b\}} &= 0 \\
%  K_{a,\{c\}} &= 0 &
  K_{a,\{b,c\}} &= 2 \\ \\
%  K_{b,\emptyset} &= 0
  K_{c,\emptyset} &= 0 &
  K_{c,\{c\}} &= 1
\end{align*}
The parameters $K_{a,\{b\}}$, $K_{a,\{c\}}$ and $K_{b,\emptyset}$
could not be inferred,
mainly due to the lack of precision in the cooperation between $b$ and $c$.
Note that the parameters $K_{b,\{b\}}$ and $K_{b,\{a,b\}}$ do not exist for this IG,
because of the absence of any self-influence $b \rightarrow b$.
\end{example}
